





\chapter*{About the Authors}

\part{Fundamentals of Deep Learning}

\chapter{Introduction to Python}
Covers key Python concepts essential for deep learning, including:
\section{Python basics (data structures, control flow, libraries).}
\section{NumPy and Pandas for data manipulation.}
\section{Matplotlib and Seaborn for visualization.}
\section{TensorFlow and PyTorch setup and basics.}

\chapter{What is Deep Learning?}
\section{Differences between traditional machine learning and deep learning.}
\section{Overview of neural networks and their capabilities.}
\section{Applications of deep learning in earth sciences.}

\chapter{Setting the Stage: The Mathematical Building Blocks of Deep Learning}
\input{chapters/Part_1/mathematics/mathematics}

\chapter{Introduction to Machine Learning}
\begin{itemize}
    \item Key ML concepts: supervised, unsupervised, and reinforcement learning.
    \item Common algorithms: linear regression, decision trees, SVMs, etc.
    \item Preparing datasets for ML.
\end{itemize}

\chapter{Getting Started with Neural Networks}
\begin{itemize}
    \item Basic structure of neural networks.
    \item Activation functions, loss functions, and optimizers.
    \item Training, validation, and evaluation of models.
\end{itemize}

\part{Deep Learning for Time Series in Water Science}

\chapter{Introduction to RNNs and LSTM}
\begin{itemize}
    \item Basics of Recurrent Neural Networks (RNNs).
    \item Long Short-Term Memory (LSTM) networks and their advantages.
    \item Applications in hydrological modeling.
\end{itemize}

\chapter{Autoencoders}
\begin{itemize}
    \item Introduction to autoencoders and their types (vanilla, sparse, denoising).
    \item Demonstrations on hydrological data.
\end{itemize}

\chapter{Sequence-to-Sequence Prediction with LSTMs}
\begin{itemize}
    \item Encoder-decoder architectures.
    \item Bidirectional LSTMs and stacked LSTMs.
    \item Real-world applications in water resource prediction.
\end{itemize}

\chapter{Attention Mechanisms}
\begin{itemize}
    \item Introduction to attention.
    \item Applications in time-series data for hydrology.
\end{itemize}

\chapter{State-of-the-Art Transformer Models for Time Series Prediction}
\begin{itemize}
    \item Introduction to transformer models.
    \item Applications in time-series prediction and generation.
\end{itemize}

\part{Deep Learning for Image Data}

\chapter{Convolutional Neural Networks (CNNs) for Image Analysis}
\begin{itemize}
    \item Basics of CNNs.
    \item Applications to remote sensing and hydrological imagery.
\end{itemize}

\chapter{LULC Classification}
\begin{itemize}
    \item Land Use Land Cover (LULC) mapping.
    \item Demonstrations using remote sensing imagery.
\end{itemize}

\chapter{Anomaly Detection in Remote Sensing Data}
\begin{itemize}
    \item Detecting anomalies in hydrological and climate data.
    \item Examples using satellite imagery.
\end{itemize}

\chapter{Digital Elevation Model (DEM) and LiDAR Data Analysis}
\begin{itemize}
    \item Processing DEM and LiDAR datasets.
    \item Applications in flood mapping and terrain analysis.
\end{itemize}

\part{Recent Advances in Deep Learning}

\chapter{Explainable AI (XAI) for Hydrological Models}
\begin{itemize}
    \item Introduction to XAI.
    \item SHAP and Integrated Gradients for feature importance.
    \item Real-world examples: drought prediction and vegetation health.
\end{itemize}

\chapter{Physics-Guided Deep Learning (PGDL)}
\begin{itemize}
    \item Combining physics-based and data-driven models.
    \item Applications in hydrology: rainfall-runoff modeling and water level prediction.
\end{itemize}










Deep Learning For Hydrology, Water Resources, Envrionmental, Climate and Earth science domains: Theory and practice with Python


Table of Contents:
Preface:
Acknowledgments:
About the book: about Who should read this book, roadmap, software reqyirements, hardwares, and Source Code examoles. 
The book is filled with varuous real world data problems and diverse priblems like: 
(1) drought classification, (2) Rainfall runoff modeling, (3) hourly temperature prediction, (4) reservior water level classification/prediction, (5) LULC classification, (6) SM prediction, (7)  heatwave classification, (8) DEM or Lidar data, (9) vegetation health detection (10) drought prediction and monitoring in real time. 
About the authors:


The book is basically divided into 3 parts broadly. 
Part I: Fundamentals of Deep Learning

Part II: Deep Learning for Time series in Water Science
part III: Deep Learning for image types like remote sensing imagery classification of images keep a small heading for this.. 

Part IV: recent advances in Deep learning like XAI or interpretable DL, Physics guided deep learning 

Part I: 

Chapter 1: Introduction to Python (make a detailed outline here we will cover few important stuffs that will help the readers in understand the python codes throughout this book)

Chapter 2: What is Deep Learning? 

Chapter 3: Setting the stage: The mathematical Building blocks of Deep Learning

Chapter 4: Introduction to Machine Learning

Chapter 5: Getting Started from the Neural Networks

Part 2: Deep learning for Time series 

Chapter 6:Introduction to RNNs and LSTM

Chapter 7: Autoencoders and demonstrate the capability of various types of AEs.

Chapter 8: Sequence to sequyence prediction with encoder decoder LSTM, biLSTM, Stacked LSTM.

Chapter 9: Attention Mechanism

Chapter 10: State-of-the-art transformer Models for time series prediction and generation

part III: DL for image from remote sensing may be or whatever title u think is good. 

basically this part will be for the remotesensing image based data. 
here we demonstrate the LULC classification, ANomaly detection, and as different types of chapters.. of DEM mapping something we will find.. 
Use of basically CNN variants. 
part IV: understanding the blackbox DL models with XAI and PGDL

Chapter 1: XAI
disucss what is XAI and we use for example the SHAP and EG methods to find somehting from the previously discusssed problems to identify the major features. 
Chapter 2: PGDL also make its outline,, 



