\chapter{Setting the Stage: Core Mathematics of Deep Learning}
\begin{mdframed}[linewidth=1pt, linecolor=gray, backgroundcolor=gray!5] \vspace{0.2cm} \textbf{\Large After finishing this chapter, you will know:} \vspace{0.1cm}

\begin{itemize}[label=\textbf{\textendash}, leftmargin=1.5em, itemsep=0.5em, labelsep=0.5em] \item \textit{How neural networks work}: Get introduced to the structure and function of neural networks through a practical example. \item \textit{The role of tensors}: Understand tensors as the core data structures for deep learning and their representations. \item \textit{Tensor operations}: Explore foundational operations like element-wise computations, broadcasting, and reshaping, which drive neural network computations. \item \textit{How neural networks learn}: Grasp the concept of backpropagation and how optimization techniques like gradient descent enable learning. \item \textit{Connecting the dots}: See how these mathematical principles come together in a real-world example. \end{itemize} \vspace{0.2cm} \end{mdframed}


Deep learning, at its core, is about building powerful models that learn to represent and predict patterns from data. However, to truly harness its potential, you need to understand the mathematical concepts that underlie it. In this chapter, we’ll gently introduce you to these foundational ideas—without overwhelming you with unnecessary mathematical jargon.

By the end of this chapter, you’ll have a strong intuition about key mathematical building blocks like tensors, operations on tensors, and gradient-based optimization techniques. We’ll walk you through practical examples, starting with a simple neural network for a drought classification task. This example will anchor our exploration of mathematical concepts such as tensors, tensor operations, gradients, and optimization.

% \begin{enumerate}
%     \item \textbf{Introducing Neural Networks Through an Example:}We begin with a hands-on example: predicting whether a region is experiencing a drought. This practical task will introduce the fundamental ideas behind neural networks and set the stage for the deeper concepts to come.


%     \item \textbf{How Neural Networks Learn: Optimization Basics:}Dive into the mechanisms that drive learning in neural networks. We’ll explore how gradients and derivatives play a critical role in adjusting the model's parameters and introduce gradient-based optimization, including the concept of backpropagation, which powers the training process.

% \item \textbf{Putting It All Together:} After building your understanding of the core principles, we’ll revisit the drought classification example to see how these concepts work in practice. This section will reinforce your learning and prepare you for applying these ideas to real-world problems in upcoming chapters.
% \end{enumerate}
\section{Introducing Neural Networks Through a Simple Example: Drought Classification}
\input{chapters/Part_1/mathematics/tensor_operations}
\input{chapters/Part_1/mathematics/optimization basics}

